---
title: "group_rep"
format: html
editor: visual
---

## Group_rep

Load dataset:

```{r}
library(tidyverse)
drugscreen <- load("~/Desktop/biol90042_group_rep/drugscreen.Rda")
gait<- load("~/Desktop/biol90042_group_rep/gait.Rda")
prs<- load("~/Desktop/biol90042_group_rep/prs.Rda")
rnaseq<- load("~/Desktop/biol90042_group_rep/rnaseq.Rda")
rnaseq_gene_info<- load("~/Desktop/biol90042_group_rep/rnaseq_gene_info.Rda")
```

### **23.3.1 Population genetics**

1.  Calculate a PRS for each participant. Which participant has the highest PRS? The lowest?

```{r}
PRS_full<-ca_co_genotypes |> 
  mutate(
    PRS=rowSums(select(ca_co_genotypes, starts_with("locus_"))*gwas_top_hits$beta
  )) #calculate the PRS for each participant
PRS_filtered<-PRS_full |> 
  select(donorid, status, starts_with("PRS"))  #select the PRS column
PRS_filtered |> 
  slice_max(PRS, n=1) #participant with the highest PRS
PRS_filtered |> 
  slice_min(PRS, n=1) #participant with the lowest PRS

```

2.  Plot and describe (with descriptive statistics) the distribution of PRS for both cases and controls.

```{r}
PRS_filtered |> 
  summarise(
    min_PRS=min(PRS),
    max_PRS=max(PRS),
    mean_PRS=mean(PRS),
    median_PRS=median(PRS),
    sd_PRS=sd(PRS),
    .by=status
  ) #descriptive statistics of the distribution of PRS for both cases and controls
#from the descriptive statistics, there is an increase of PRS in cases group compare to the controls group, where the max PRS remain similar
ggplot(
  data=PRS_filtered,
  aes(x=PRS, fill=status))+
  geom_histogram(bins=15)+ #creating a histogram that range into 15 bars, choose 15 is because it is much smoother compare to bins=30 which is the default, and easier to see overall distribution
  facet_wrap(~status)+ #make two separate plots based on the status
  labs(
    title="Distribution of PRS for both cases and controls",
    x="Polygenic risk score",
    y="Count",
    )+
  theme_minimal()+ #to minimize background, look better
  theme(legend.position="none") #because each panel is already labelled with case and control, so having the legend is meaningless, therefore it is removed

```

3.  Is there a statistically significant difference in the PRS of cases and controls? Ensure you check all the assumptions of the test you choose.

```{r}
#check normality visually
ggplot(PRS_filtered, aes(sample=PRS))+
  geom_qq()+
  geom_qq_line()+
  facet_wrap(~status)
#QQ-plot does suggest normality for case, but do not suggest normality for ctrl, confirm normality using Shapiro Wilks test
PRS_filtered |> 
  filter(status=="case") |> 
  pull(PRS) |> 
  shapiro.test()
PRS_filtered |> 
  filter(status=="ctrl") |> 
  pull(PRS) |> 
  shapiro.test()
#for case, the p value is 0.1856, no evidence against normality
#for ctrl, the p value is 0.0006253, very strong evidence against normality
#to do a t-test, both group need to meet the normality assumption, since the ctrl group has violated the assumption, t-test is not appropriate here. Wilcox test will be applied here to compare the two groups.
wilcox.test(PRS~status, data=PRS_filtered)
#the wilcoxon test statistics value is 16028 with a p-value of 2.005e-10. This suggests a extremely strong evidence that the PRS distribution differ between the Pakinson disease case group and the control group, with case group have higher PRS compare to the control group
```

### **23.3.2 RNA-seq of patient iPSC-derived neurons**

Firstly, load the packages

```{r}
library(edgeR)
library(RNAseq123)
library(limma)
library(pheatmap)
library(ggrepel)
```

1.  Which genes are differentially expressed between PD cases and healthy controls? Please also visualise your DE results with appropriate plot(s).

```{r}
#Firstly, we need to filter our lowly expressed genes to increase our statistical power. To do this, we will use the filterByExpr() function in the edgeR package. This will keep genes with a count of 10 or more in at least 8 samples (as we have 8 samples per group)
keep <- filterByExpr(dge)
#To check the proportion of genes we have ommited, we can create a proportions table:
proportions(table(keep)) #We are keeping 97.77% of our genes
#to apply this filter to the DGE object, we will use [] to filter out FALSE hits (lowly expressed genes) and assign this dge
dge <- dge[keep, , keep.lib.sizes = FALSE]

#Now we need to normalise our data by converting it to counts per million (CPM), so we can compare samples and groups, and account for RNA composition bias. To do this we will use the TMM method to calculate the normalisation factor for each sample. normLibSizes() function can be used to calculate normalisation factors
dge <- normLibSizes(dge, method = "TMM")
#the normalisation factors in dge$samples are close to 1 in all samples, suggesting minimal difference in RNA composition between samples

#Now we will fit a linear model using the limma-voom framework by first creating a design matrix to represent covariates

design <- model.matrix(~0 + group, data = dge$samples)

#We will first need to rename our design matrix to be cleaner
colnames(design) <- c("Control", "Case")

#Contrast matrix
contr_matrix <- makeContrasts(
  CasevsControl = "Case - Control",
  levels = design
)

v <- voom(dge, design, plot = TRUE)
vfit <- lmFit(v)
#apply constrasts
vfit <- contrasts.fit(vfit, contrasts = contr_matrix)
#statistical test
efit <- eBayes(vfit)
#decide tests function to decide if a gene is differentially expressed or not
dt <- decideTests(efit)
summary(dt)
top <- topTable(efit, coef = "CasevsControl", number = Inf)
#converted to tibble to make it easier to work with
top_tibble <- as_tibble(top)


#MA plot
limma::plotMA(efit, coef = "CasevsControl", status = dt[, "CasevsControl"])

##ggplot MA plot
#first need to assign gene names to the efit$genes dataframe

#turning efit$genes into character
efit$genes$genes <- as.character(efit$genes$genes)
#so I can now change efit$genes to include SYMBOL column which matches the entrez_genename_key#entrezid to efit$genes$genes, and supplies the entrez_genename_key$gene_name to the efit$genes dataframe
efit$genes$SYMBOL <- entrez_genename_key$gene_name[
  match(efit$genes$genes, entrez_genename_key$entrezid)]

#now organise fata from efit into a tibble format
ma_plot_data <- tibble(logFC = efit$coefficients[, "CasevsControl"],
                       AvgExpr = efit$Amean,
                       status = factor(as.numeric(dt[, "CasevsControl"])),
                       gene = efit$genes$SYMBOL,
                       p_value = efit$p.value[, "CasevsControl"])

ma_plot_data_fct <- ma_plot_data %>%
  mutate(
    # replace status with a new factor
    status = factor(
      status,
      # levels (order is not DE, down, up)
      levels = c(0, -1, 1),
      # labels for the numbers so they are readble
      labels = c("Not DE", "Down", "Up")
    )
  )
#labelling of top most significant genes
top_de_genes <- ma_plot_data_fct %>%
  slice_min(n = 15, p_value)
#colours to label points by
scale_cols <- c(
  "Not DE" = "black",
  "Down" = "#1a0099",
  "Up" = "#990000"
)
#MA ggplot
ggplot(ma_plot_data_fct, aes(x = AvgExpr, y = logFC, col = status)) +
  geom_point(data = ma_plot_data_fct %>% filter(status == "Not DE"), size = 0.1) +
  geom_point(data = ma_plot_data_fct %>% filter(status != "Not DE"), size = 1) +
  geom_label_repel(
    data = top_de_genes,
    aes(label = gene, col = NULL),
    max.overlaps = 20,
    show.legend = FALSE
  ) +
  scale_colour_manual(values = scale_cols) + 
  theme_classic() +
  labs(title = "MA plot for Case vs Control")

#Volcano Plot
ggplot(ma_plot_data_fct, aes(x = logFC, y = -log10(p_value), col = status)) +
  geom_point(data = ma_plot_data_fct %>% filter(status == "Not DE"), size = 0.1) +
  geom_point(data = ma_plot_data_fct %>% filter(status != "Not DE"), size = 1) +
  geom_label_repel(
    data = top_de_genes,
    aes(label = gene, col = NULL),
    max.overlaps = 20,
    show.legend = FALSE
  ) +
  scale_colour_manual(values = scale_cols) + 
  theme_classic() +
  labs(title = "Volcano plot for Case Vs Control")
```

2.  Which gene ontology families do the DE genes belong to?

```{r}

```

3.  Is the clustering of samples by gene expression consistent with their disease status as cases and controls?

```{r}
#Now we want to check that our samples group together as we'd expect based on our experimental design, by making an MDS plot. This will show us whether or not the controls group together, and the PD patients group together. To do this, we will use the plotMDS() function and then make the MDS plot in ggplot2
plotMDS(dge, labels = dge$samples$group)

mds_result <- plotMDS(dge)
#create tibble
mds_tibble <- tibble(
  x = mds_result$x,
  y = mds_result$y,
  sample = colnames(dge),
  group = dge$samples$group
)
#
dim1_var_explained <- round(mds_result$var.explained[1] * 100)
dim2_var_explained <- round(mds_result$var.explained[2] * 100)
#create ggplot
ggplot(data = mds_tibble,
       aes(x = x, y = y, colour = group)) + geom_point() + geom_text_repel(aes(label = sample)) +
  labs(
    x = paste0("Leading logFC dim 1 (", dim1_var_explained, "%)"),
    y = paste0("Leading logFC dim 2 (", dim2_var_explained, "%)")
  )
#IMPROVE PLOT
# From this plot we can see that 5/8 of the PD patient groups cluster to the left of the map, while the other 3 PD patients cluster with the 8 control cases
```

### **23.3.3 Drug screening**

1.  Using the `geom_tile()` function, visualise the luminescence across the plate. Do there appear to be any spatial patterns in this data?

```{r}
# Convert the plate data to long format, with each row representing one well.
plate1_long <- plate1 |> 
  pivot_longer(
    cols = contains("col"),
    names_to = "column",
    values_to = "RLU"
  ) |> 
  # create numeric row and column coordinates for plotting
  mutate(
    row_num = as.numeric(str_remove(plate_row, "row")),
    col_num = as.numeric(str_remove(column, "col"))
  )

# Heat map
ggplot(plate1_long, aes(x = col_num, y = row_num, fill = RLU)) +
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkblue", limits = c(429, 742), breaks = seq(400, 750, by = 50)) +
  labs(
    title = "Luminescence across the 1536-well plate",
    x = "Column",
    y = "Row",
    fill = "Luminescence (RLU)"
  ) +
  theme_minimal()

# Most values have relatively similar luminescence values.
# A few individual wells have much higher luminescence, coloured with dark blue.
# No obvious consistent gradient from one side of the plate to the other. 
# No obvious edge effect where wells around the edges are consistently higher or lower.
```

2.  Which compounds are statistically significantly associated with increased neuronal viability after MPP exposure? Ensure that you check your assumptions.

```{r}
# Convert the plate_meta data to long format, with each row representing one well.
plate_meta_long <- plate_meta |> 
  pivot_longer(
    cols = contains("col"),
    names_to = "column",
    values_to = "compound"
  )

# Combine the luminescence results with compound information based on each well's row and column.
drug_data <- plate_meta_long |> 
  left_join(
    plate1_long,
    by = c("plate_row", "column")
  )

# Separate the compound ID from the test/control label at the last underscore.
drug_data <- drug_data |> 
  mutate(
    group = str_extract(compound, "test|ctrl"),
    compound = str_remove(compound, "_(test|ctrl)$")
  )

# Summarise luminescence by test and control groups.
compound_summary <- drug_data |> 
  summarise (
    mean_RLU = mean(RLU),
    median_RLU = median(RLU),
    sd_RLU = sd(RLU),
    n = n(),
    .by = c(compound, group)
  ) |> 
  pivot_wider(
    names_from = group,
    values_from = c(mean_RLU, median_RLU, sd_RLU)
  ) |> # Put test and control results side by side for each compound.
  mutate(
    difference = mean_RLU_test - mean_RLU_ctrl
  ) # Calculate the difference in mean RLU between test and control groups.

# Welch's t-test assumption checking
# First, check the normality of RLU for some compounds
drug_data |> 
  filter(compound %in% c( "cmpd_6406", "cmpd_4834", "cmpd_6255")) |> 
  ggplot(aes(sample = RLU)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~compound + group)
# The QQ plots show no obvious deviation from normality across these compounds. However, normality is difficult to confirm with only four replicates per group.

#Perform Welch's t-test without assuming equal variances.
t_test_results <- drug_data |> 
  summarise(
    p_value = t.test(RLU ~group, var.equal = FALSE)$p.value,
    .by = compound
  )
# Adjust p-values for multiple testing using the Benjamini-Hochberg method.
t_test_results <- t_test_results |> 
  mutate(
    p_adjusted = p.adjust(p_value, method = "BH")
  )

# Combine t-test results with compound summary information
t_test_results <- t_test_results |> 
  left_join(
    compound_summary |> 
      dplyr::select(compound, difference), # Keep only the compound ID and mean RLU difference.
    by = "compound"
  )
# Find significant compounds associated with increased neuronal viability.
t_test_results |> 
  filter(p_adjusted < 0.05,
         difference > 0)

# Conclusion: Four compounds (cmpd_5386, cmpd_5118, cmpd_7466 and cmpd_6972) significantly increased neuronal viability after MPP exposure.
```

3.  What are the mechanisms of action and targets of the compounds you identify in Q2?

```{r}
# Find the target and mechanism of action for the significant compounds.
compound_dict |> 
  filter(id %in% c("cmpd_5386", "cmpd_5118", "cmpd_7466", "cmpd_6972"))

# NOT SURE IF BELOW IS NECESSARY:
# Combine significant compounds with their targets and mechanisms of action.
question3_results <- t_test_results |> 
  filter(p_adjusted < 0.05, difference > 0) |> 
  left_join(
    compound_dict,
    by = c("compound" = "id")
  )
```

### **23.3.4 Clinical trial**

1.  Plot each participants’ gait data and describe it using descriptive statistics. Which of these descriptive statistics do you think is most useful for capturing the overall gait smoothness and balance of a participant?

```{r}

```

2.  Is there a statistically significant difference in gait between participants receiving L-DOPA and the new compound? Ensure that you check your assumptions.

```{r}

```

3.  Does either of age or the exercise regime affect gait? If so, take this into consideration when investigating the relationship between the treatment and gait.

```{r}

```

### **23.3.5 Bringing it together/linking between datasets**

1.  Using their unique donor IDs, how many patients appear across all four arms of the study?

```{r}

```

2.  Do any variables other than gene expression distinguish the RNA-seq cases from controls? (Hint: try colouring points on your MDS plot by other information you have about the donors, from the other datasets)

```{r}

```

3.  Are any of the gene targets of hit compounds from the drug screen, differentially expressed in RNA-seq? Which GO terms are they associated with and what might this tell you about the pathology of PD?

```{r}

```

4.  Is there any relationship between the PRS and gait of patients in the clinical trial?

```{r}

```

5.  In a paragraph, summarise your findings and suggest directions for future research.

```{r}

```
