---
title: "group_rep"
format: html
editor: visual
---

## Group_rep

Load dataset:

```{r}
library(tidyverse)
drugscreen <- load("../group3_data/drugscreen.Rda")
gait<- load("../group3_data/gait.Rda")
prs<- load("../group3_data/prs.Rda")
rnaseq<- load("../group3_data/rnaseq.Rda")
rnaseq_gene_info<- load("../group3_data/rnaseq_gene_info.Rda")
```

### **23.3.1 Population genetics**

1.  Calculate a PRS for each participant. Which participant has the highest PRS? The lowest?

```{r}
PRS_full<-ca_co_genotypes |> 
  mutate(
    PRS=rowSums(select(ca_co_genotypes, starts_with("locus_"))*gwas_top_hits$beta
  )) #calculate the PRS for each participant
PRS_filtered<-PRS_full |> 
  select(donorid, status, starts_with("PRS"))  #select the PRS column
PRS_filtered |> 
  slice_max(PRS, n=1) #participant with the highest PRS
PRS_filtered |> 
  slice_min(PRS, n=1) #participant with the lowest PRS

```

2.  Plot and describe (with descriptive statistics) the distribution of PRS for both cases and controls.

```{r}
PRS_filtered |> 
  summarise(
    min_PRS=min(PRS),
    max_PRS=max(PRS),
    mean_PRS=mean(PRS),
    median_PRS=median(PRS),
    sd_PRS=sd(PRS),
    .by=status
  ) #descriptive statistics of the distribution of PRS for both cases and controls
#from the descriptive statistics, there is an increase of PRS in cases group compare to the controls group, where the max PRS remain similar
ggplot(
  data=PRS_filtered,
  aes(x=PRS, fill=status))+
  geom_histogram(bins=15)+ #creating a histogram that range into 15 bars, choose 15 is because it is much smoother compare to bins=30 which is the default, and easier to see overall distribution
  facet_wrap(~status)+ #make two separate plots based on the status
  labs(
    title="Distribution of PRS for both cases and controls",
    x="Polygenic risk score",
    y="Count",
    )+
  theme_minimal()+ #to minimize background, look better
  theme(legend.position="none") #because each panel is already labelled with case and control, so having the legend is meaningless, therefore it is removed

```

3.  Is there a statistically significant difference in the PRS of cases and controls? Ensure you check all the assumptions of the test you choose.

```{r}
#check normality visually
ggplot(PRS_filtered, aes(sample=PRS))+
  geom_qq()+
  geom_qq_line()+
  facet_wrap(~status)
#QQ-plot does suggest normality for case, but do not suggest normality for ctrl, confirm normality using Shapiro Wilks test
PRS_filtered |> 
  filter(status=="case") |> 
  pull(PRS) |> 
  shapiro.test()
PRS_filtered |> 
  filter(status=="ctrl") |> 
  pull(PRS) |> 
  shapiro.test()
#for case, the p value is 0.1856, no evidence against normality
#for ctrl, the p value is 0.0006253, very strong evidence against normality
#to do a t-test, both group need to meet the normality assumption, since the ctrl group has violated the assumption, t-test is not appropriate here. Wilcox test will be applied here to compare the two groups.
wilcox.test(PRS~status, data=PRS_filtered)
#the wilcoxon test statistics value is 16028 with a p-value of 2.005e-10. This suggests a extremely strong evidence that the PRS distribution differ between the Pakinson disease case group and the control group, with case group have higher PRS compare to the control group
```

### **23.3.2 RNA-seq of patient iPSC-derived neurons**

Firstly, load the packages

```{r}
library(edgeR)
library(RNAseq123)
library(limma)
library(pheatmap)
library(ggrepel)
```

1.  Which genes are differentially expressed between PD cases and healthy controls? Please also visualise your DE results with appropriate plot(s).

```{r}

```

2.  Which gene ontology families do the DE genes belong to?

```{r}

```

3.  Is the clustering of samples by gene expression consistent with their disease status as cases and controls?

```{r}

```

### **23.3.3 Drug screening**

1.  Using the `geom_tile()` function, visualise the luminescence across the plate. Do there appear to be any spatial patterns in this data?

```{r}

```

2.  Which compounds are statistically significantly associated with increased neuronal viability after MPP exposure? Ensure that you check your assumptions.

```{r}

```

3.  What are the mechanisms of action and targets of the compounds you identify in Q2?

```{r}

```

### **23.3.4 Clinical trial**

1.  Plot each participants’ gait data and describe it using descriptive statistics. Which of these descriptive statistics do you think is most useful for capturing the overall gait smoothness and balance of a participant?

```{r}

```

2.  Is there a statistically significant difference in gait between participants receiving L-DOPA and the new compound? Ensure that you check your assumptions.

```{r}

```

3.  Does either of age or the exercise regime affect gait? If so, take this into consideration when investigating the relationship between the treatment and gait.

```{r}

```

### **23.3.5 Bringing it together/linking between datasets**

1.  Using their unique donor IDs, how many patients appear across all four arms of the study?

```{r}

```

2.  Do any variables other than gene expression distinguish the RNA-seq cases from controls? (Hint: try colouring points on your MDS plot by other information you have about the donors, from the other datasets)

```{r}

```

3.  Are any of the gene targets of hit compounds from the drug screen, differentially expressed in RNA-seq? Which GO terms are they associated with and what might this tell you about the pathology of PD?

```{r}

```

4.  Is there any relationship between the PRS and gait of patients in the clinical trial?

```{r}

```

5.  In a paragraph, summarise your findings and suggest directions for future research.

```{r}

```
